Minimizing Time Required For Graduation

### 2010-12-05 paper/paper.tex
fiddled with paper type and margin settings slightly
proof read, chopped out a couple sentences, fixed some awkward ones, added in a couple places

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margins=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

@@

\section{Introduction} \subsection{Background} Registration can be a stressful
time for every student at the University of Washington. Without deliberate
planning, students must decide which classes to take on a quarterly basis. The
decision process involves considering factors such as quarterly class
offerings, the amount of credits, the prerequisites of the classes, time
conflicts, and how the choices in current quarter affect future quarters. The
University of Washington registration system provides a schedule finder;
however, it only takes into account time conflicts, and does not give advice on
how to pick the most ideal schedule. The situation described above makes it
difficult for students advising faculties. A student's graduation time may be
extended because of his or her inability to take a class that has limited
offerings during that particular year, for reasons such as not satisfying the
prerequisites of the class. The advising office’s workload can also be increased
due to the changing plans and majors of students and their need for repeated
advising.

@@

\subsection{Importance} The implementation of the algorithm as part of the
school’s registration system will benefit the students by enabling them to map
out their quarterly class schedules until graduation, and to preview their ideal
schedule, the length of time until graduation, and the credit hours per quarter.
This will lead to better decision making for what majors to pursue, and given
the student's current state, which schedules are the most ideal in terms of
saving time and resources. This will also help students save money, as they will
be in school for fewer quarters. The university also benefits from the reduced
labor cost and increased efficiency of the advising offices.

@@

Every quarter the Math department offers a subset of classes from all of the
available courses, each class is offered during a certain subset of available
time periods. Since there are far more classes than time periods during the
regular five-day week there are inevitably conflicts between courses that
a student wants to take and courses that a student needs to take. Or worse,
there are conflicts between two or more courses that a students needs to
graduate. To graduate with a B.S. in Mathematics there are certain classes that
are required and then there are elective requirements too. Additionally
a student may wish to take some classes that are not required.  Adding to the
problem is many classes cannot just be taken whenever they are available;
instead, certain prerequisites must be satisfied first. In the case of math
courses, there can be multiple different prerequisite sets that satisfy the
requirements for a class.

@@

We must now also define some variables used throughout our paper.
\begin{itemize} \item Let $S$ be the set of courses required to be taken in
order to graduate. $S$ can be thought of as our goal state, and can also contain
elective courses a student simply wishes to take.  \item Let $T$ be the set of
courses a given student has already completed. If no courses have been taken,
this set may be empty.  \item Let $U$ be the set of all courses a student could
possibly ever take.  \item Let $P$ be a set of rules specifying course
prerequisites.  \item Let $l$ be the maximum number of credits a student wishes
to take during any given quarter.  \item Let $O$ be the set of all 50-minute
time blocks between 8:30am and 7:30pm from Monday through Friday.  \end{itemize}

\subsection{Representing Prerequisites} Initially, we hoped to represent course
prerequisites using a directed acyclic graph, in which each node was a course,
and edges between nodes $i$ and $j$ if and only if $i$ was a prerequisite for
$j$.  This method was used by Pritsker et. al. and Xiangtong et. al. However, it
does not work in our model, and courses can have a complicated set of multiple
prerequisites. For example, {\it MATH327} requires both $MATH126$ and $MATH300$,
or just $MATH136$. Let this graph be called $G_{prereq}$. Due to these
complications, a $G_{prereq}$ alone will not suffice.

@@

\subsection{Initial Model and Tractability} With a static schedule it would be
possible to deterministically plan an entire schedule from start to finish,
unfortunately we have neither the computational resources nor the luxury of
a static schedule. From the student’s point of view, although there are some
classes that are offered every quarter, senior electives are not, and sometimes
it cannot be predicted more than one quarter in advance when a senior elective
will be offered. Even given perfect information, where we know the possible
class schedules for every quarter we face another problem; finding the best
possible schedule reduces to enumerating possible class schedules for every
quarter until the graduation requirements are satisfied. Given that we cannot
predict in advance the conflicts between classes, we cannot even determine the
branching factor of the enumeration tree nor its depth. Since we can only
determine one quarter in advance, this suggests approaching the problem by using
heuristics to pick the best possible classes each quarter.

Instead of trying to solve the complete problem we chose to use the rolling
horizon approach used by Xiangtong et al \cite{xiangton:informs} when scheduling
classes for airlines. Our problem already has a well defined horizon
demarcation, the quarters. For the rest of the paper we focus on developing
methods for attempting to optimally pick classes in each quarter separately.

@@

\subsection{Determining Available Course Offerings} We need to select the
classes we can currently take and then build conflict graphs for classes based
on the times they are offered. In this section, we let $L$ be the initially
empty set of course offerings a given student is eligible to take in the given
quarter, $q$. \\

@@

Since we are dealing with asymptotic running time we will drop the constant
terms and combining all results yields an overall run time of $O(n)$.

\subsection{Selection of Quarterly Classes} \subsubsection{Selection Procedure}
At this point in time, we know which course offerings are available for the
given student to register for, and we must now pick the most optimal ones. There
a several ways we could view the problem of determining which classes to take
each quarter. We will initially encode the class selection as a binary integer
programming problem in spirit of the methods presented by Dinkel et al
\cite{dinkel:scheduling}. and Pritsker et al \cite{prisker:informs}. At this
point we have the set $L$ containing possible classes we that we need to take.

\subsubsection{Heuristic Value Function} In order to determine which classes
should be picked in a given schedule, we use a heuristic function which assigns
each course offering a value. The higher the value of this function for a given
course offering is, the more likely the given course offering will be placed in
the schedule. Recalling that our goal is to minimize the total number of
quarters taken (and implicitly, maximize the total number of courses taken per
quarter), our heuristic function should clearly give higher value to course
offerings that will help us achieve this goal. One component to the heuristic
function then must be the number of dependent courses a given course offering
has. If a course satisfies many prerequisites, then we know that there are many
courses that we cannot complete without it, so we want our heuristic function to
favor it.

@@

\subsubsection{Formulation as Binary Integer Program} We can now formulate our
problem as a binary integer linear programming problem. Before stating the
linear program, we must first define some variables used in it: \\
\begin{itemize} \item $ X_i = \left\{ \begin{array}{lr} 1 \text{ if we take
} X_i\\ 0 \text{ if we don't take } X_i \end{array} \right. $ \item For all $t
\in O$, we let $t = 1$.  \item Let $m = \|L\|$.  \end{itemize} We now state our
problem as follows:\\ Maximize: $$ \text{Value of Quarter} = \sum_{i=1}^m
Value(c_i) * X_i $$ Subject To: $$ \sum_{i=1}^m Weight(c_i) * X_i \leq l, $$ $$
\forall t_j \in O, \sum_{i=1}^m t_j * X_i \leq 1, $$ $$ X_i \in \{0, 1\} $$ The
linear program can be interpreted as follows: \begin{itemize} \item The value of
a quarter is the total number of courses taken that quarter.  \item In each
quarter, the total number of credits cannot exceed the specified limit.  \item
Only one course can be taken in one particular time slot \item A course can only
be taken or not taken. In other words, decision variables are binary.  \item
Each non-zero decision variable in the optimal solution represents a class we
will take.  \end{itemize}

>>>>

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{top=3cm,bottom=3cm,left=3cm,right=3cm,letterpaper} % or letterpaper (US) or a5paper or....

@@

\section{Introduction} \subsection{Background} Registration can be a stressful
time for every student at the University of Washington. Without deliberate
planning, students must decide which classes to take on a quarterly basis. The
decision process involves considering factors such as quarterly class
offerings, the amount of credits, the prerequisites of the classes, time
conflicts, and how the choices in current quarter affect future quarters. The
University of Washington registration system provides a schedule finder;
however, it only takes into account time conflicts, and does not give advice on
how to pick the most ideal schedule.

A student's graduation time may be extended because of his or her inability to
take a class that has limited offerings during that particular year.
Additionally, the advising office’s workload can also be increased due to the
changing plans and majors of students and their need for repeated advising.

@@

\subsection{Importance} The implementation of the algorithm as part of the
school’s registration system will benefit the students by enabling them to map
out their quarterly class schedules until graduation. Students will be able to
preview their ideal schedule, the length of time until graduation, and the
credit hours per quarter.  This will lead to better decision making for what
majors to pursue, and given the student's current state, which schedules are the
most ideal in terms of saving time and resources. This will also help students
save money, as they will be in school for fewer quarters. The university also
benefits from the reduced labor cost and increased efficiency of the advising
offices.

@@

Every quarter the Math department offers a subset of classes from all of the
available courses, each class is offered during a certain subset of available
time periods. Since there are far more classes than time periods during the
regular five-day week there are inevitably time conflicts between courses that
a student could take. To graduate with a B.S. in Mathematics there are certain
classes that are required and then there are elective requirements too. Adding
to the problem is many classes cannot just be taken whenever they are available;
instead, certain prerequisites must be satisfied first. In the case of math
courses, there can be multiple different prerequisite sets that satisfy the
requirements for a class.

@@

Additionally: \begin{itemize} \item Let $S$ be the set of courses required to be
taken in order to graduate. $S$ can be thought of as our goal state, and can
also contain elective courses a student simply wishes to take.  \item Let $T$ be
the set of courses a given student has already completed. If no courses have
been taken, this set may be empty.  \item Let $U$ be the set of all courses
a student could possibly ever take.  \item Let $P$ be a set of rules specifying
course prerequisites.  \item Let $l$ be the maximum number of credits a student
wishes to take during any given quarter.  \item Let $O$ be the set of all
50-minute time blocks between 8:30am and 7:30pm from Monday through Friday.
\end{itemize}

\subsection{Representing Prerequisites} Initially, we hoped to represent course
prerequisites using a directed acyclic graph. Each node was a course, and edges
existed between nodes $i$ and $j$ if and only if $i$ was a prerequisite for $j$.
However, it does not work in our model, and courses can have a complicated set
of multiple prerequisites.  For example, {\it MATH327} requires both $MATH126$
and $MATH300$, or just $MATH136$.

@@

\subsection{Initial Model and Tractability} With a static schedule it would be
possible to deterministically plan an entire schedule from start to finish,
unfortunately we have neither the computational resources nor the luxury of
a static schedule to do plan. From the student’s point of view, although there
are some classes that are offered every quarter, senior electives are not, and
sometimes it cannot be predicted more than one quarter in advance when a senior
elective will be offered. Even given perfect information, where we know the
possible class schedules for every quarter we face another problem; finding the
best possible schedule reduces to enumerating possible class schedules for every
quarter until the graduation requirements are satisfied. Given that we cannot
predict in advance the conflicts between classes, we cannot even determine the
branching factor of the enumeration tree nor its depth. Since we can only
determine one quarter in advance, this suggests approaching the problem by using
heuristics to pick the best possible classes each quarter.

Instead of trying to solve the complete problem we chose to use the rolling
horizon approach used by Xiangtong et al \cite{xiangton:informs} when scheduling
classes for airlines. Our problem already has a well defined horizon
demarcation, the quarters. For the rest of the paper we focus on developing
methods to optimally pick classes in each quarter separately.

@@

\subsection{Determining Available Course Offerings} We need to select the
classes we can currently take and then build conflict graphs for classes based
on the times they are offered. In this section, we let $L$ be the initially
empty set of course offerings a given student is eligible to take in the given
quarter, $q$.

@@

Since we are dealing with asymptotic running time we will drop the constant
terms and combining all results yields an overall runtime of $O(n)$.

\subsection{Selection of Quarterly Classes} \subsubsection{Selection Procedure}
At this point in time, we know which course offerings are available for the
given student to register for, and we must now pick the most optimal ones.  We
present two ways of approaching quarterly class choice. We will initially encode
the class selection as a binary integer programming problem in spirit of the
methods presented by Dinkel et al \cite{dinkel:scheduling}. and Pritsker et al
\cite{prisker:informs}. A second approach developed by Yamada et al
\cite{yamada:heuristic} and Pferschy et al \cite{pferschy:kcg} is presented for
dealing with very large class choices where binary LP would be difficult solve.

\subsubsection{Heuristic Value Function} In order to determine which classes
should be picked in a given schedule, we use a heuristic function which assigns
each course offering a value. The higher the value of this function for a given
course offering is, the more likely the given course offering will be placed in
the schedule. Recalling that our goal is to minimize the total number of
quarters taken (and implicitly, maximize the total number of courses taken per
quarter). Our heuristic function should give higher value to course offerings
that will help us achieve this goal. One component to the heuristic function
then must be the number of dependent courses a given course offering has. If
a course satisfies many prerequisites, then we know that there are many courses
that we cannot complete without it, so we want our heuristic function to favor
it.

@@

\subsubsection{Formulation as Binary Integer Program} We can now formulate our
problem as a binary linear programming problem. Before stating the linear
program, we must first define some variables used in it: \\ \begin{itemize}
\item $ X_i = \left\{ \begin{array}{lr} 1 \text{ if we take } X_i\\ 0 \text{ if
we don't take } X_i \end{array} \right. $ \item For all $t \in O$, we let $t
= 1$.  \item Let $m = \|L\|$.  \end{itemize} We now state our problem as
follows:\\ Maximize: $$ \text{Value of Quarter} = \sum_{i=1}^m Value(c_i) * X_i
$$ Subject To: $$ \sum_{i=1}^m Weight(c_i) * X_i \leq l, $$ $$ \forall t_j \in
O, \sum_{i=1}^m t_j * X_i \leq 1, $$ $$ X_i \in \{0, 1\} $$ The linear program
can be interpreted as follows: \begin{itemize} \item The value of a quarter is
the total number of courses taken that quarter.  \item In each quarter, the
total number of credits cannot exceed the specified limit.  \item Only one
course can be taken in one particular time slot \item A course can only be taken
or not taken. In other words, decision variables are binary.  \item Each
non-zero decision variable in the optimal solution represents a class we will
take.  \end{itemize}
